Investigating Determinism of Game Engines for Autonomous Vehicle Verification

### 2020-04-27 paper/cav_determinisim_main.tex
move results to investigation section

% ***************************************************
%  CASE STUDY
% ***************************************************
\section{Empirical Investigation} \label{s:case-study}
We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}. The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance %$\sigma^2$
for any actor should be less than a user-defined threshold.

@@

\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles. The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding or avoiding with other actors. These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.

The tests listed in Table~\ref{TableOfExperiments} were chosen to cover a range of interactions between actor types. Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2). In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller. In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point. In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback in the game engine, see Fig~\ref{Test_a}. Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors. Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.

@@

\subsection{Experiment Description}\label{s:results}
\todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. Repeated traces are compared at the same time index to provide a value for the variance of actor $a$ with respect to time $t$, $\sigma_a(t)$. The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
\begin{equation}
\max_{a,t}\sigma_a(t), 
\end{equation}
herein referred to as $\max\sigma$ for brevity.
% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
The $\max\sigma$ value was analysed against the different scenarios and settings that were identified as potential sources of simulation variance, and compared against the \textit{deterministic threshold} to indicate if the simulation was deterministic or not. 
%
% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}

@@

\subsection{Internal Settings}

@@

An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings. The results were compared to a baseline of the default engine settings. 
%
However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values. Details on this previous investigation can be found on the Trustworthy Systems github\footnote{\url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}.

\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}

After discovering that internal engine settings could not achieve a simulation variance that met the required \textit{deterministic threshold}, settings external to the game engine were explored. 
%
Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment. 
%
For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}. This hypothesis was supported by our initial work performed on the unreal engine\footnote{see footnote 14} and was explored more fully in this work.
%
To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation. 
%
% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively. Resource utilisation figures reported here should be considered approximate values.
% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).

\subsection{System Configuration and Screening}
These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz. To ensure reliability of results, each test was repeated 1000 times. Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$. To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system. These were:

@@

% ***************************************************
%  RESULTS
% ***************************************************
\subsection{Results and Discussion}\label{ResultsSection}
The summary of all the tests are shown in the final two columns of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels. From these  results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels. However, for other scenarios involving vehicles or a mixture of actors types, the resource utilisation level has a significant impact on \textit{simulation variance}.

Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU). In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description.} A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s).}

However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation. Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision. This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism. \todo{Why ``not reliable''.}

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation in this configuration.}
\end{figure*}

>>>>

% ***************************************************
%  Empirical Investigation
% ***************************************************
\section{Empirical Investigation} \label{s:case-study}
We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}. The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance %$\sigma^2$
for any actor should be less than a user-defined threshold.

@@

% ***************************************************
%  Scenario Description
% ***************************************************
\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles. 
%
These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.
%
The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.

% ***************************************************
%  Actor Collisions
% ***************************************************
\subsection{Actor Collisions}\label{S:Actor_Collisions}
Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Collision/Overview/index.html}}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}. 
%
Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.

The tests are listed in Table~\ref{TableOfExperiments} and were chosen to cover a range of interactions between actor types. 
%
Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2). In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller. In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point. In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback in the game engine, see Fig~\ref{Test_a}. Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors. Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.

@@

% ***************************************************
%  Experiment Description
% ***************************************************
\subsection{Experiment Description}\label{s:results}
\todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. Repeated traces are compared at the same time index to provide a value for the variance of actor $a$ with respect to time $t$, $\sigma_a(t)$. The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
\begin{equation} \label{eq:max_sigma}
\max_{a,t}\sigma_a(t), 
\end{equation}
herein referred to as $\max\sigma$ for brevity.
% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
The $\max\sigma$ value was analysed against the different scenarios and settings that were identified as potential sources of simulation variance, and compared against the \textit{deterministic threshold} to indicate if the simulation was deterministic or not. 
%
% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}

@@

% ***************************************************
%  Internal Settings
% ***************************************************
\subsection{Internal Settings}

@@

An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings. The results were compared to a baseline of the default engine settings. 
%
However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values. Details on this previous investigation can be found on the Trustworthy Systems github~\cite{TSLUnrealEngineTesting}.

% ***************************************************
%  External Settings
% ***************************************************
\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}

After discovering that internal engine settings could not achieve a simulation variance that met the required \textit{deterministic threshold}, settings external to the game engine were explored. 
%
Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment. 
%
For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}. This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the Carla platform.

To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation. 
%
% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively. Resource utilisation figures reported here should be considered approximate values.
% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{System Configuration and Screening}
These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz. To ensure reliability of results, each test was repeated 1000 times. Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$. To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system. These were:

@@

% ***************************************************
%  Results and Discussion
% ***************************************************
\subsection{Results and Discussion}\label{ResultsSection}
A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted) where the value reported is the maximum across all resource utilisation levels, i.e. the worst case for a given scenario.  
%
From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) are those that are consistently deterministic over all resource utilisation levels with or without a collision. However, for the other scenarios involving vehicles or a mixture of actors types, the all breach the deterministic threshold that has been set, with some deviation in actor path as large as 59cm.

Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
%
Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU). In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
%
\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
%
A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (see Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed $cm$ threshold), some are consistently below the threshold (test ID 5 \& 6) and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).

\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation time.
%
Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision. This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism. \todo{Why ``not reliable''.}

@@
